---
output:
  revealjs::revealjs_presentation:
    incremental: false
    self_contained: true
    center: true
    theme: default
    css: custom.css
    transition: slide
    slide_level: 1
    reveal_options:
      width: 100%
      height: 100%
    includes:
      in_header: header.html
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height=5)
```


# Visualization in R:{data-background-color="#4E2A84" .white-font}

## <span class="white-font">`ggplot2` and `plotly`</span>
\
\
Christina Maimone  



# Workshop Outline

Visualization in R \
`ggplot`: Examples and Background \
What is Plotly?\
Let's make some plots \


# Visualization in R {data-background-color="#E4E0EE" .gray-font}

---

## Built-in Base R Graphics

R comes with plotting functions built in.  

---

```{r}
hist(rnorm(100))
```

---

```{r}
plot(x=runif(30), y=runif(30))
```

---

* Good for a lot of situations
* But they often aren't very attractive
* Grouping is difficult
* Legends can be tricky and aren't automatic
* Different plots are made in different ways

---

## R Plotting Systems

* base graphics
* `ggplot2`
* `lattice`: around for a long time, not widely used
* `ggvis`: builds on `ggplot2`, extending for interactive graphics, but not complete yet

---

## Additional Visualization Packages

* `plotly`
* `igraph`: for network graphs
* `ggmap` and `leaflet`: maps (other packages for this too)
* `googleVis`: Google Charts API
* `rgl`: 3D plots
* `dygraphs`: interactive time series

\
Other specialized packages for fields such as genomics

---

# ggplot {data-background-color="#E4E0EE" .gray-font}

---

Let's start with the same plots as before

```{r}
library(ggplot2)
```

---

```{r, error=FALSE, warning=FALSE, message=FALSE}
mydata<-data.frame(vals=rnorm(100)) ## ggplot expects a data.frame
ggplot(mydata, aes(vals))+geom_histogram()
```


---

```{r}
mydata<-data.frame(x=runif(30), y=runif(30))
ggplot(mydata, aes(x, y))+geom_point()
```

---

Those look a little better (maybe)...

\
But the code is more complicated

---

## Could use the `qplot` shorcut 

\
`qplot` is a `ggplot2` function

\
Works a lot the base graphics `plot`

---

```{r, message=FALSE}
qplot(rnorm(100))
```

---

```{r}
qplot(runif(30), runif(30))
```

---

But we're going to ignore `qplot` and only use `ggplot` to avoid confusion

---

## Better Examples

`ggplot` is most useful for more complicated plots

---

Basic Scatterplot

```{r}
# using mpg data set that comes with ggplot2
ggplot(mpg, aes(displ, hwy)) + geom_point() 
```

---

Add a third dimension with color

```{r}
ggplot(mpg, aes(displ, hwy, colour = class)) + geom_point() 
```

---

Add linear trend lines by class

```{r}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point() + geom_smooth(method="lm", se=FALSE)
```

---

Separate by class

```{r}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point() + geom_smooth(method="lm", se=FALSE) + 
  facet_wrap(~class)
```

---

## A few polished examples

---

# {data-background-image="http://minimaxir.com/img/ggplot-tutorial/buzzfeed-listicle-scatterplot.png" data-background-size=contain class="full"}

<div class="caption"><a href="http://minimaxir.com/2015/02/ggplot-tutorial/">Source: Max Woolf</a></div>

---

# {data-background-image="https://raw.githubusercontent.com/minimaxir/movie-gender/76581d9468d4b59dfbd6963c21a6c85e2339bd82/movie-gender-10.png" data-background-size=contain class="full"}

<div class="caption"><a href="https://github.com/minimaxir/movie-gender/blob/master/movie_gender.ipynb">Source: Max Woolf</a></div>

---

# {data-background-image="http://juliasilge.com/figs/2017-04-15-Gender-Pronouns/austen-1.png" data-background-size=contain class="full"}

<div class="caption"><a href="http://juliasilge.com/blog/Gender-Pronouns/">Source: Julia Silge</a></div>




---

## `ggplot` or `ggplot2`?

The name of the package is <span style="color:#836EAA">`ggplot2`</span>.  

The name of the function used to create a plot is <span style="color:#836EAA">`ggplot`</span>.  

---

## But...

You'll see `ggplot` used to refer to the whole system as well (as I do here).  `ggplot2` started with `ggplot`, which isn't used anymore.

---

## What makes up a `ggplot` plot?

---

## Data

<span style="color:#836EAA">`ggplot(mpg)`</span>

\
a `data.frame` to use

---

## Aesthetics

`ggplot(mpg,` <span style="color:#836EAA">`aes(x=displ, y=hwy, color=class)`</span>`)`

\
position (x, y) \
size &bull; color &bull; shape \
fill &bull; linetype &bull; transparency

\
<small>See https://cran.r-project.org/web/packages/ggplot2/vignettes/ggplot2-specs.html</small>

---

## Geoms

`ggplot(mpg, aes(x=displ, y=hwy, color=class)) + `\
<span style="color:#836EAA">`geom_point()`</span>

\
What type of plot to draw?

One or two variables?

Discrete or continuous values?

---

## Other Components

* **Coordinates**: the grid you're plotting on
* **Stats**: transformations of the data
* **Scales**: breaks, colors, labels, transformations
* **Facets**: plots for different groups
* **Labels**: title, axes
* **Themes**: overall style


---


# Plotly {data-background-color="#E4E0EE" .gray-font}

---

## Use Plotly for

* Interactive graphics
* Sharing plots online
* Editable plots (zoom, select observations, etc.)
* Teams working in multiple languages (MATLAB, R, Python, JavaScript)

---

# {data-background-image="https://cpsievert.github.io/plotly_book/images/gapminder-highlight-animation.gif" data-background-size=contain class="full"}

<div class="caption"><a href="https://cpsievert.github.io/plotly_book/">Source: Carson Sievert</a></div>

---

https://plot.ly/~RPlotBot/3260.embed



# Code {data-background-color="#4E2A84" .white-font}

This presentation and the code we're using are available at https://www.github.com/nuitrcs/ggplotworkshop/







